Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016

### 2019-04-21 draft-paper.Rmd
fixed the margins for the paper and revised introduction

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremyforbes1995@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation is employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna"
wpnumber:
jelcodes: C31, C33, D72
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from Australian federal elections.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas federal elections, conducted by the Australian Electoral Commission (AEC), usually occur every three years or so. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections --- election years in which a Census does not occur. The predominant approach in previous studies was to join voting outcomes to the nearest Census, without accounting for any temporal differences [see @DavisStimson98;@Stimson06;@Liao09;@Stimson09]. Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate "Census-like" data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \@ref(data). All data and associated documentation relating to this procedure are available in the `eechidna` `R` package [@eechidna], providing a resource for future analysis.

Previous work on modelling Australian federal elections has found that aggregate socio-demographics are relatively good predictors of voting outcomes. @Forrest01 used multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. @Stimson06, @Stimson09 and @Stimson12 used principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. In contrast, @Stimson09 models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of the swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional dataset. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from previous work.

The cross-sectional dataset for each election used here consists of the two-party preferred vote (as the response variable), and a set of common socio-demographic variables (as the explanatory variables) that characterize each electorate. To prepare these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into "factors". To account for the inherent spatial structure of the data, a spatial error model is then estimated for each election. These steps are discussed in Section \@ref(modelling). In Section \@ref(results) inference is conducted to see which effects are significant and how these effects change over time.

Section \@ref(conclusion) concludes, followed by two supplementary sections.

>>>>

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremyforbes1995@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation is employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna"
wpnumber:
jelcodes: C31, C33, D72
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time.

Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from Australian federal elections. Joining these two data sources is problematic as there is an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas federal elections, conducted by the Australian Electoral Commission (AEC), usually occur every three years or so. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections --- election years in which a Census does not occur. The predominant approach in previous studies was to join voting outcomes to the nearest Census, without accounting for any temporal differences [see @DavisStimson98;@Stimson06;@Liao09;@Stimson09]. Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate "Census-like" data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \@ref(data). All data and associated documentation relating to this procedure are available in the `eechidna` `R` package [@eechidna], providing a resource for future analysis.

Previous work on modelling Australian federal elections has found that aggregate socio-demographics are relatively good predictors of voting outcomes. @Forrest01 used multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. @Stimson06, @Stimson09 and @Stimson12 used principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. In contrast, @Stimson09 models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of the swing vote can be explained by Census data.

Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed. Each federal election is modelled with a cross-sectional dataset. <!--The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from previous work.-->
The cross-sectional dataset for each election used here consists of the two-party preferred vote (as the response variable), and a set of common socio-demographic variables (as the explanatory variables) that characterize each electorate. To prepare these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into "factors". To account for the inherent spatial structure of the data, a spatial error model is then estimated for each election.

The paper is organised as follows. Section \@ref(data) describes the data collection, joining and cleaning. These pre-processing steps and model details are discussed in Section \@ref(modelling). Section \@ref(results) describes the inference conducted to determine signficance of effects and how these change over time. Section \@ref(conclusion) summarises the work. Two supplementary sections document the contributions of others to this work and the software.

### 2019-04-21 paper.Rmd
fixed the margins for the paper and revised introduction

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author: Jeremy Forbes$^*$, Dianne Cook, Rob J Hyndman
toc: false
documentclass: anzsauth
classoption: times, doublespace
fontfamily: times
bibliography: references.bib
biblio-style: anzsj
output: 
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: anzjs_preamble.tex
---

@@

**Keywords:** federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna

@@

hume %>%
  kable(longtable=FALSE, booktabs=TRUE, digits=4,
        format='latex',escape=FALSE,
        col.names=c(
          "Source zone (2016)",
          "Proportion",
          "Source zone population",
          "Population allocated to purple region: $\\hat{P}_{s,t}$",
          "AusCitizen (\\%): $D_s$"),
    caption = "Population from each intersecting Census source zone (2016) that is allocated to the target zone (purple region - Hume electoral boundary in 2013), and the corresponding proportion of Australian citizens in each of these source zones.") %>%
  kableExtra::kable_styling(font_size=10, latex_options='scale_down')
```

>>>>

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author: Jeremy Forbes$^*$, Dianne Cook, Rob J Hyndman
toc: false
documentclass: anzsauth
classoption: times, doublespace
fontfamily: times
bibliography: references.bib
biblio-style: anzsj
geometry: "left=5cm,top=5cm"
output: 
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: anzjs_preamble.tex
---

@@

**Keywords:** federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna

@@

hume %>%
  kable(longtable=FALSE, booktabs=TRUE, digits=4,
        format='latex',escape=FALSE,
        col.names=c(
          "Source zone (2016)",
          "Proportion",
          "Source zone population",
          "Population allocated to purple region: $\\hat{P}_{s,t}$",
          "AusCitizen (\\%): $D_s$"),
    caption = "Population from each intersecting Census source zone (2016) that is allocated to the target zone (purple region - Hume electoral boundary in 2013), and the corresponding proportion of Australian citizens in each of these source zones.") %>%
  kableExtra::kable_styling(font_size=12, latex_options='scale_down')
```

### 2019-04-21 paper.tex
fixed the margins for the paper and revised introduction

\documentclass[times, doublespace]{anzsauth}
\usepackage[]{times}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016},
            pdfauthor={Jeremy Forbes\^{}*, Dianne Cook, Rob J Hyndman},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}\vspace{0.5cm}}
\preauthor{\begin{center}\large}
\postauthor{\end{center}\begin{center}\emph{Department of Econometrics \& Business Statistics, Monash University, Australia}\footnotetext{$^*$Corresponding author. Email: \texttt{jeremyforbes1995@gmail.com}}\end{center}}

@@

\textbf{Keywords:} federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna

@@

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3\% and 53.5\% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from Australian federal elections.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas federal elections, conducted by the Australian Electoral Commission (AEC), usually occur every three years or so. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections --- election years in which a Census does not occur. The predominant approach in previous studies was to join voting outcomes to the nearest Census, without accounting for any temporal differences (see Davis and Stimson 1998; Stimson, McCrea, and Shyy 2006; Liao, Shyy, and Stimson 2009; Stimson and Shyy 2009). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate ``Census-like'' data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation (Goodchild, Anselin, and Deichmann 1993) and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \ref{data}. All data and associated documentation relating to this procedure are available in the \texttt{eechidna} \texttt{R} package (Forbes et al. 2019), providing a resource for future analysis.

Previous work on modelling Australian federal elections has found that aggregate socio-demographics are relatively good predictors of voting outcomes. Forrest et al. (2001) used multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. Stimson, McCrea, and Shyy (2006), Stimson and Shyy (2009) and Stimson and Shyy (2012) used principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. In contrast, Stimson and Shyy (2009) models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of the swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional dataset. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from previous work.

The cross-sectional dataset for each election used here consists of the two-party preferred vote (as the response variable), and a set of common socio-demographic variables (as the explanatory variables) that characterize each electorate. To prepare these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into ``factors''. To account for the inherent spatial structure of the data, a spatial error model is then estimated for each election. These steps are discussed in Section \ref{modelling}. In Section \ref{results} inference is conducted to see which effects are significant and how these effects change over time.

Section \ref{conclusion} concludes, followed by two supplementary sections.

@@

\caption{\label{tab:hume1}Population from each intersecting Census source zone (2016) that is allocated to the target zone (purple region - Hume electoral boundary in 2013), and the corresponding proportion of Australian citizens in each of these source zones.}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}{lrrrr}
\toprule
Source zone (2016) & Proportion & Source zone population & Population allocated to purple region: $\hat{P}_{s,t}$ & AusCitizen (\%): $D_s$\\
\midrule
Hume & 0.9654 & 150643 & 145427 & 90.0168\\
Riverina & 0.2511 & 155793 & 39117 & 89.1144\\
Eden-Monaro & 0.1109 & 147532 & 16358 & 87.9999\\
Canberra & 0.0028 & 196037 & 548 & 85.4793\\
Fenner & 0.0023 & 202955 & 474 & 83.6432\\
\addlinespace
Whitlam & 0.0006 & 152280 & 92 & 89.5173\\
Gilmore & 0.0006 & 150436 & 86 & 89.0266\\
Calare & 0.0001 & 161298 & 21 & 87.5603\\
\bottomrule
\end{tabular}}
\end{table}

@@

\leavevmode\hypertarget{ref-Forrest01}{}%
Forrest, James, Margaret Alston, Chris Medlin, and Siti Amri. 2001. ``Voter Behaviour in Rural Areas: A Study of the Farrer Electoral Division in Southern New South Wales at the 1998 Federal Election.'' \emph{Australian Geographical Studies} 39 (2). Wiley/Blackwell: 167--82.

\leavevmode\hypertarget{ref-Goodchild1993}{}%
Goodchild, M F, L Anselin, and U Deichmann. 1993. ``A Framework for the Areal Interpolation of Socioeconomic Data.'' \emph{Environment and Planning A} 25 (3). London, England: SAGE Publications: 383--97.

@@

\leavevmode\hypertarget{ref-Stimson09}{}%
Stimson, Robert, and Tung-Kai Shyy. 2009. ``A Socio-Spatial Analysis of Voting for Political Parties at the 2007 Federal Election.'' \emph{People and Place} 17 (1). Monash University, Centre for Population; Urban Research: 39--54.

>>>>

\documentclass[times, doublespace]{anzsauth}
\usepackage[]{times}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[left=5cm,top=5cm]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016},
            pdfauthor={Jeremy Forbes\^{}*, Dianne Cook, Rob J Hyndman},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}\vspace{0.5cm}}
\preauthor{\begin{center}\large}
\postauthor{\end{center}\begin{center}\emph{Department of Econometrics \& Business Statistics, Monash University, Australia}\footnotetext{$^*$Corresponding author. Email: \texttt{jeremyforbes1995@gmail.com}}\end{center}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\textbf{Keywords:} federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna

@@

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3\% and 53.5\% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time.

Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from Australian federal elections. Joining these two data sources is problematic as there is an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas federal elections, conducted by the Australian Electoral Commission (AEC), usually occur every three years or so. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections --- election years in which a Census does not occur. The predominant approach in previous studies was to join voting outcomes to the nearest Census, without accounting for any temporal differences (see Davis and Stimson 1998; Stimson, McCrea, and Shyy 2006; Liao, Shyy, and Stimson 2009; Stimson and Shyy 2009). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate ``Census-like'' data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation (Goodchild, Anselin, and Deichmann 1993) and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \ref{data}. All data and associated documentation relating to this procedure are available in the \texttt{eechidna} \texttt{R} package (Forbes et al. 2019), providing a resource for future analysis.

Previous work on modelling Australian federal elections has found that aggregate socio-demographics are relatively good predictors of voting outcomes. Forrest et al. (2001) used multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. Stimson, McCrea, and Shyy (2006), Stimson and Shyy (2009) and Stimson and Shyy (2012) used principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. In contrast, Stimson and Shyy (2009) models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of the swing vote can be explained by Census data.

Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed. Each federal election is modelled with a cross-sectional dataset.
The cross-sectional dataset for each election used here consists of the two-party preferred vote (as the response variable), and a set of common socio-demographic variables (as the explanatory variables) that characterize each electorate. To prepare these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into ``factors''. To account for the inherent spatial structure of the data, a spatial error model is then estimated for each election.

The paper is organised as follows. Section \ref{data} describes the data collection, joining and cleaning. These pre-processing steps and model details are discussed in Section \ref{modelling}. Section \ref{results} describes the inference conducted to determine signficance of effects and how these change over time. Section \ref{conclusion} summarises the work. Two supplementary sections document the contributions of others to this work and the software.

@@

\caption{\label{tab:hume1}Population from each intersecting Census source zone (2016) that is allocated to the target zone (purple region - Hume electoral boundary in 2013), and the corresponding proportion of Australian citizens in each of these source zones.}
\centering
\resizebox{\linewidth}{!}{
\fontsize{12}{14}\selectfont
\begin{tabular}{lrrrr}
\toprule
Source zone (2016) & Proportion & Source zone population & Population allocated to purple region: $\hat{P}_{s,t}$ & AusCitizen (\%): $D_s$\\
\midrule
Hume & 0.9654 & 150643 & 145427 & 90.0168\\
Riverina & 0.2511 & 155793 & 39117 & 89.1144\\
Eden-Monaro & 0.1109 & 147532 & 16358 & 87.9999\\
Canberra & 0.0028 & 196037 & 548 & 85.4793\\
Fenner & 0.0023 & 202955 & 474 & 83.6432\\
\addlinespace
Whitlam & 0.0006 & 152280 & 92 & 89.5173\\
Gilmore & 0.0006 & 150436 & 86 & 89.0266\\
Calare & 0.0001 & 161298 & 21 & 87.5603\\
\bottomrule
\end{tabular}}
\end{table}

@@

\leavevmode\hypertarget{ref-Forrest01}{}%
Forrest, James, Margaret Alston, Chris Medlin, and Siti Amri. 2001. ``Voter Behaviour in Rural Areas: A Study of the Farrer Electoral Division in Southern New South Wales at the 1998 Federal Election.'' \emph{Australian Geographical Studies} 39 (2): 167--82.

\leavevmode\hypertarget{ref-Goodchild1993}{}%
Goodchild, M F, L Anselin, and U Deichmann. 1993. ``A Framework for the Areal Interpolation of Socioeconomic Data.'' \emph{Environment and Planning A} 25 (3): 383--97.

@@

\leavevmode\hypertarget{ref-Stimson09}{}%
Stimson, Robert, and Tung-Kai Shyy. 2009. ``A Socio-Spatial Analysis of Voting for Political Parties at the 2007 Federal Election.'' \emph{People and Place} 17 (1): 39--54.
